The impacts of park access on health outcomes: A spatial comprehensive approach

### 2019-02-28 manuscript.Rnw
Update with reviews from Nico and Kari

\corrauth{Gregory Macfarlane\\
430 Engineering Building\\
Brigham Young University\\
Provo, UT 84602\\
USA}
\begin{abstract}
Recent research has underscored the potential for public green spaces to
influence individual and societal health outcomes, but
empirical measurements of such influences have yielded mixed results to date, with
particular disagreement surrounding how access to parks ought to be defined while
controlling alternate explanations. In this paper we present a comprehensive
measure of park accessibility drawn from random utility choice theory, which avoids
arbitrary assertions of proximity while incorporating potentially numerous
amenities and attributes of both the parks and the population. We apply this
choice-based accessibility measure to Census tract-level health and
socioeconomic data in New York City, paired with geographic parks data and
geolocated position data collected from the Twitter API. Controlling for the
socioeconomic variables and spatially correlated error terms, we show a positive
but suggestive relationship between park access and physical activity rates, and
a clear negative relationship between park access and obesity rates.
\end{abstract}

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
According to \citet{Duhl1999} however, the field of public health has
moved toward a new paradigm that is grounded in a social model of health,
treating health ``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.''
Further, \citet{Coutts2008} maintains that awareness of the impacts of the built
environment on health will increase connections between the practices of public
health and urban planning. As professionals from both fields begin to operate
from the assumption that the design and configuration of the built environment
can facilitate or inhibit physical activity, they are increasingly looking to
public spaces like parks and greenways as key elements of the built environment
that can support physical activity \citep{Bedimo-Rung2005a}.

In spite of this interest and the numerous studies attempting to empirically
estimate the strength of such support, evidence to date has been mixed.
A major challenge researchers have faced is a proliferation of techniques for
measuring access to green spaces, of which many are plausible but lack
analytical or theoretical basis. In this paper, we present a holistic
and flexible measurement for park accessibility based in random utility
choice theory. This choice-based accessibility measure compares the continuous
distance to all parks in the region, weighted against the size of the park and
its other amenities. We apply this measurement to study the link between park
accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatial correlation and tract-level socioeconomic characteristics.
We find a positive relationship where the least park-accessible tracts have an
expected physical activity participation rate roughly a full percentage point
lower than the most accessible tracts; a traditional quarter-mile buffer
analysis estimates an effect twice as large but marginally less significant.
However, we also find a strong and significant correlation between
the choice-based park accessibility metric and decreased obesity rates --- again
controlling for spatial correlation, socioeconomic attributes, and physical
activity rates. A traditional buffer analysis does not show significance.

@@

In a study of neighborhoods in New York City, \citet{Stark2014} used
hierarchical linear models to explore relationships between the proportion of
zip codes that was park space, the cleanliness of parks, and the body mass index
of adult residents. The authors found that a
higher proportion of park space in a neighborhood was associated with a lower
BMI for its residents, and lower park cleanliness scores were associated with a
higher BMI, suggesting that increasing the supply of clean, well-maintained
parks could positively impact adult BMI.
\citet{Suminski2005} showed that adding vegetation to
existing sidewalks and trails increased local motivation to engage in physical
activity.

@@

Note that this formulation is an extension of the gravity-based accessibility
statistics, in the same way that the gravity-based trip distribution model
is a specific case of
a destination-choice model \citep{Daly1982}. But the extension is meaningful, and
allows us to consider more attributes of the park or the individual beyond size
and impedance. For example, we can aggregate the number of tweets emanating from
a park as a measure of the park's attractiveness beyond its size, and the
utility for the choice maker from Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d + tweets_j\lambda_t
\end{equation}

@@

The Centers for Disease Control and Prevention makes small-area estimates on
key health indicators available through its 500 Cities data program. The models
are estimated on Census sociodemographic data paired with national and local
health surveys, and have been validated at multiple geographic levels
\citep{Wang2018, Wang2017, Zhang2015}. We use two synthetic indicators as our
outcome variables in this study: the share of adults in a Census tract who are
obese, and the share of adults who participate in no leisure-time physical activity.
To improve clarity in our interpretation, we use the complement of the second variable ---
the share of tract adults who participate in {\em some} physical activity ---
even if the amount is not sufficient to affect overall health. Both indicators
are estimated for the year 2016.

@@

To the health data, we join data collected through the Census Bureau API via
the {\texttt tidycensus} package for R \citep{Walker2019}. The primary dataset
is a geographic polygons layer of Census tracts in the five boroughs of New York
City. We append to each Census tract relevant sociodemographic data
for each tract level from the American Community Survey 2013-2017 5-year
estimates. For a small handful of tracts in our sample, Census supressed the
median income estimate; these appear to be primarily wealthy tracts and in
almost all cases the CDC estimates of obesity and physical activity are missing
as well. After removing these tracts from the estimation dataset, we have
\Sexpr{nrow(tracts)} complete cases. Table \ref{tab:tractsdata} presents key
descriptive statistics for these data.

@@

To this base model we add a measure of park accessibility considered in two
ways: a logsum-based statistic as described earlier and a more typical quarter-mile
binary access buffer. The estimated coefficients for these models are also
given in Table \ref{tab:pa_models}. In both accessibility specifications,
the access measure is positively correlated with physical activity rates; all else
equal, people in tracts with higher accessibility to parks have higher estimated
rates of physical activity. Though the estimated coefficient on the quarter-mile
buffer appears to have a larger absolute value than the logsum-based statistic,
this masks a key difference in the specification worth remembering.
The logsum allow for variation in access unavailable in the buffer-based method.
In our dataset almost every tract
(\Sexpr{round(sum(tracts$access_025) / nrow(tracts) * 100, 0)} \%) is within
a quarter-mile of {\em some} kind of park, but this park may not be large enough
to support physical activities all individuals would like to participate in.
Because tracts with only adequate park access are grouped with tracts having
excellent park access, the overall effect is comparatively unclear. The logsum-based
coefficient, on the other hand, suggest going from the least-accessible ($-3.0$)
to the most-accessible ($3.0$) tract implies roughly a $1.1$\% percentage-point
rise in estimated physical activity rates

@@

As discussed when presenting the choice-based accessibility metric above,
a key benefit it offers is the natural way in which park amenities beyond simply size
can be accomodated. For example, the number of tweets emanating from a park
might be seen as a measure of the park's popularity, or a proxy for its use. It
may be construed that parks with high popularity or use contain other amenities
that encourage residents to use the park space, raising physical activity rates
and lowering obesity rates beyond what would be expected with the park's size
and proximity alone. It is reasonable to question whether park size and Twitter
activity are highly correlated, and thus would result in a double-counting of
park size in the accessibility statistic. In our data the number of tweets is
positively but only loosely correlated  with park
size ($\rho = \Sexpr{round(cor(parks$log_size, parks$log_tweets), 3)}$),
indicating that the Twitter data in fact could provide new information or proxy
for the desired amenities.

@@

Figure \ref{fig:tweets_map} shows the percent change in the normalized log
sum statistic after including the twitter infromation. There is not a recognizable
overarching pattern to the changes, though tracts immediately surrounding
Prospect Park and Central Park gain even more relative accessibility, and tracts
in midtown Manhattan, Queens, and southern Bronx tend to lose some.

@@

Table \ref{tab:tweets_coefs} presents the estimated coefficients relating
accessibility to obesity and physical activity rates, with and without the
inclusion of Twitter data. In both cases, including twitter information in
the accessibility statistic tempers the strength of the relationship with the
dependent variable and modestly widens the standard error without substantively
affecting the significance of the test statistics or the interpretation.
From this limited example, we cannot say whether Twitter activity enhances the
ability of parks to induce physical activity rates or lower obesity rates,
though further investigation of this and other park amenities would certainly
be warranted.

@@

We readily acknowledge limitations in this study. As in any study conducted 
with areal data, we at risk of falling victim to the ecological inference 
fallacy, where aggregate statistics mask or contradict disaggregated or 
individual-level trends. A large-sample survey of individuals in New York City,
including measured physical activities and obesity would always be prefferable
to the tract-level data used in this study. It would be also be
preferrable to have obtained the $\vect{\lambda}$ accessibility utility coefficients 
from a high-quality survey of leisure destination choice rather than
asserting them to match our prior expectations of what constitutes quality
park access. An ideal survey to address the question would incorporate both sets
of questions: physical activity and health data on one hand and park use (including
which parks were used and how frequently) on the other. As no such dataset
exists to our knowledge, this tract-level aggregate analysis with asserted 
utility coefficients is the possibility that remains.

This paper presents a holistic accessibility statistic that could, in theory,
accommodate many attributes of the destination parks as well as the people who 
might use them. As an illustration: the park-going population could be separated
into at least four delineated clusters, each preferring different amenities of a 
park: 
\begin{itemize}
 \item{runners and cyclists: long, interesting trail systems}
 \item{sports players: soccer fields, basketball courts, or baseball diamonds}
 \item{families with small children: water features and playgrounds}
 \item{casual users: water features, gardens, performances, etc.}
\end{itemize} 
An analyst could then compute the accessibility logsum for each cluster with
different utility values for each park's amenities, and obtain a measure of
a neighborhood's accessibility to park features that its residents most care 
about. In this paper, we proceed only incrementally beyond this theory by 
adding Twitter activity as an an element of a park's attractiveness above and
beyond its size. Exploring additional amenities or market segmentation strategies
could provide a more definitive understanding of the relationship between park 
accessibility and health outcomes.

As a travel impedance measure, we used the Euclidean distance between each tract's 
population-weighted centroid and the nearest point on the edge of a park's border.
Euclidean distances have well-rehearsed limitations regarding their fidelity with 
the underlying infrastructure network, etc.. Network-based distances can also
suffer from challenges when applied to multimodal problems; these challenges are
exacerbated when the non-highway mode share is high, as is likely when considering
access to parks. A better metric of travel impedance may be a mode choice model
log-sum, which weights all travel alternatives against each other. A

Finally, this study is primarily focused on the hypothesis that accessibility to
parks encourages physical activity, which in turn reduces obesity. There are a 
multitude of other hypotheses that might be proposed and tested with the basic
methodology we have employed here, or competing explanations for the outcomes we
have observed. It is distinctly possible, for instance, that individuals who wish
to exercise regularly in parks choose to live near them; this could potentially
explain why the observed effects of accessibility seem partially dependent on
the representation of income in our econometric model. Controlling for this
potential self-selection effect would be necessary to isolate the exogenous
impacts of park access on obesity or other health outcomes. And regarding these
other variables: this study did not consider potential relationships between park
access and hospitalization rates, life expectancy, respiratory disease, 
mental health, or any number of potential beneficial outcomes hypothesized 
or explored in the existing literature. Exploring these connections and their
underlying mechanisms should be a priority as city planners and urban architects
attempt to improve the quality of life of urban residents in the future.

\begin{funding}
This project was funded in part via \ldots.
\end{funding}

>>>>

\corrauth{Gregory Macfarlane\\
430 Engineering Building\\
Brigham Young University\\
Provo, UT 84602\\
USA}
\begin{abstract}
Recent research has underscored the potential for public green spaces to
influence individual and societal health outcomes, but
empirical measurements of such influences have yielded mixed results to date, with
particular disagreement surrounding how access to parks ought to be defined while
controlling alternate explanations. In this paper we present a comprehensive
measure of park accessibility drawn from random utility choice theory, which avoids
arbitrary assertions of proximity while incorporating potentially numerous
amenities and attributes of both the parks and the population. We apply this
choice-based accessibility measure to Census tract-level health and
socioeconomic data in New York City, paired with geographic parks data and
geolocated position data collected from the Twitter API. Controlling for the
socioeconomic variables and spatially correlated error terms, we show a positive
and significant relationship between park access and physical activity rates, and
a clear negative relationship between park access and obesity rates. Applying
a more traditional analysis considering 1/4 mile acess to parks yields mixed
results.
\end{abstract}

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
According to \citet{Duhl1999} however, the field of public health has
moved toward a new paradigm that is grounded in a social model of health,
treating health ``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.'' (p. 7)
Further, \citet{Coutts2008} maintains that awareness of the impacts of the built
environment on health will increase connections between the practices of public
health and urban planning. As professionals from both fields begin to operate
from the assumption that the design and configuration of the built environment
can facilitate or inhibit physical activity, they are increasingly looking to
public spaces like parks and greenways as key elements of the built environment
that can support physical activity \citep{Bedimo-Rung2005a}.

In spite of this interest and the numerous studies attempting to empirically
estimate the strength of such support, evidence to date has been mixed.
A major challenge researchers have faced is a proliferation of techniques for
measuring access to green spaces, of which many are plausible but lack
analytical or theoretical basis. In this paper, we present a holistic
and flexible measurement for park accessibility based in random utility
choice theory. This choice-based accessibility measure compares the continuous
distance to all parks in the region, weighted against the size of the park and
its other amenities.

We apply this measurement to study the link between park
accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatial correlation and tract-level socioeconomic characteristics.
We find a positive relationship where the least park-accessible tracts have an
expected physical activity participation rate roughly a full percentage point
lower than the most accessible tracts; a traditional quarter-mile buffer
analysis estimates an similar albeit marginally less significant effect.
However, we also find a strong and significant correlation between
the choice-based park accessibility metric and decreased obesity rates --- again
controlling for spatial correlation, socioeconomic attributes, and physical
activity rates. In this case a traditional buffer analysis does not show
significance.

@@

In a study of neighborhoods in New York City, \citet{Stark2014} used
hierarchical linear models to explore relationships between the proportion of
zip codes that was park space, the cleanliness of parks, and the body mass index
of adult residents. The authors found that a
higher proportion of park space in a neighborhood was associated with a lower
BMI for its residents, and lower park cleanliness scores were associated with a
higher BMI, suggesting that increasing the supply of clean, well-maintained
parks could positively impact adult BMI.

@@

Note that this formulation is an extension of the gravity-based accessibility
statistics, in the same way that the gravity-based trip distribution model
is a specific case of
a destination-choice model \citep{Daly1982}. But the extension is meaningful and
allows us to consider more attributes of the park or the individual beyond size
and impedance. For example, we can aggregate the number of tweets emanating from
a park as a measure of the park's attractiveness beyond its size, and the
utility for the choice maker from Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d + tweets_j\lambda_t
\end{equation}

@@

The Centers for Disease Control and Prevention makes small-area estimates on
key health indicators available through its 500 Cities data program \citep{CDC5002016}. The models
are estimated on Census sociodemographic data paired with national and local
health surveys, and have been validated at multiple geographic levels
\citep{Wang2018, Wang2017, Zhang2015}. We use two synthetic indicators as our
outcome variables in this study: the share of adults in a Census tract who are
obese, and the share of adults who participate in no leisure-time physical activity.
To improve clarity in our interpretation, we use the complement of the second variable ---
the share of tract adults who participate in {\em some} physical activity ---
even if the amount is not sufficient to affect overall health. Both indicators
are estimated for the year 2016.

@@

To the health data, we join data collected through the Census Bureau API via
the \texttt{tidycensus} package for R \citep{Walker2019}. The primary dataset
is a geographic polygons layer of Census tracts in the five boroughs of New York
City. We append to each Census tract relevant sociodemographic data
for each tract from the American Community Survey 2013-2017 5-year
estimates. For a small handful of tracts in our sample, Census supressed the
median income estimate; these appear to be primarily wealthy tracts and in
almost all cases the CDC estimates of obesity and physical activity are missing
as well. After removing these tracts from the estimation dataset, we have
\Sexpr{nrow(tracts)} complete cases. Table \ref{tab:tractsdata} presents key
descriptive statistics for these data.

@@

To this base model we add a measure of park accessibility considered in two
ways: a logsum-based statistic as described earlier and a more typical quarter-mile
binary access buffer. The estimated coefficients for these models are also
given in Table \ref{tab:pa_models}. In both accessibility specifications,
the access measure is positively correlated with physical activity rates; all else
equal, people in tracts with higher accessibility to parks have higher estimated
rates of physical activity. Though the estimated coefficient on the quarter-mile
buffer appears to have a larger absolute value than the logsum-based statistic,
this masks a key difference in the specification worth further comment.
The logsum allow for variation in access unavailable in the buffer-based method.
In our dataset almost every tract
(\Sexpr{round(sum(tracts$access_025) / nrow(tracts) * 100, 0)} \%) is within
a quarter-mile of {\em some} kind of park, but this park may not be large enough
to support physical activities all individuals would like to participate in.
Because tracts with only adequate park access are grouped with tracts having
excellent park access, the overall effect is comparatively unclear. The logsum-based
coefficient, on the other hand, suggests going from the least-accessible ($-3.0$)
to the most-accessible ($3.0$) tract implies roughly a $1.1$\% percentage-point
rise in estimated physical activity rates

@@

As discussed when presenting the choice-based accessibility metric above,
a key benefit it offers is the natural way in which park amenities beyond simply size
can be accommodated. For example, the number of tweets emanating from a park
might be seen as a measure of the park's popularity, or a proxy for its use. It
may be construed that parks with high popularity or use contain other amenities
that encourage residents to use the park space, raising physical activity rates
and lowering obesity rates beyond what would be expected with the park's size
and proximity alone. It is reasonable to question whether park size and Twitter
activity are highly correlated, and thus would result in a double-counting of
park size in the accessibility statistic. In our data the number of tweets is
positively but only loosely correlated  with park
size ($\rho = \Sexpr{round(cor(parks$log_size, parks$log_tweets), 3)}$),
indicating that the Twitter data in fact could provide new information or proxy
for the desired amenities.

@@

Figure \ref{fig:tweets_map} shows the percent change in the normalized log
sum statistic after including the twitter information. There is not a recognizable
overarching pattern to the changes, though tracts immediately surrounding
Prospect Park and Central Park gain even more relative accessibility, and tracts
in midtown Manhattan, Queens, and southern Bronx tend to lose some.

@@

Table \ref{tab:tweets_coefs} presents the estimated coefficients relating
accessibility to obesity and physical activity rates, with and without the
inclusion of Twitter data. In both cases, including twitter information in
the accessibility statistic tempers the strength of the relationship with the
dependent variable and modestly widens the standard error without substantively
affecting the significance of the test statistics or the interpretation.
From this limited example, we cannot say whether Twitter activity enhances the
ability of parks to increase physical activity rates or lower obesity rates,
though further investigation of this and other park amenities would certainly
be warranted.

@@

We readily acknowledge limitations in this study. As in any study conducted
with areal data, we are at risk of falling victim to the ecological inference
fallacy, where aggregate statistics mask or contradict disaggregated or
individual-level trends. A large-sample survey of individuals in New York City,
including measured physical activities and obesity would always be preferable
to the tract-level data used in this study. It would also be
preferrable to have obtained the $\vect{\lambda}$ accessibility utility coefficients
from a high-quality survey of leisure destination choice rather than
asserting them to match our prior expectations of what constitutes quality
park access. An ideal survey to address the question would incorporate both sets
of questions: physical activity and health data on one hand and park use (including
which parks were used and how frequently) on the other. As no such dataset
exists to our knowledge, this tract-level aggregate analysis with asserted
utility coefficients is the possibility that remains.

This paper presents a holistic accessibility statistic that could, in theory,
accommodate many attributes of the destination parks as well as the people who
might use them. As an illustration: the park-going population could be separated
into at least four delineated clusters, each preferring different amenities of a
park:
\begin{itemize}
 \item{runners and cyclists: long, interesting trail systems}
 \item{sports players: soccer fields, basketball courts, or baseball diamonds}
 \item{families with small children: water features and playgrounds}
 \item{casual users: water features, gardens, performances, etc.}
\end{itemize}
An analyst could then compute the accessibility logsum for each cluster with
different utility values for each park's amenities, and obtain a measure of
a neighborhood's accessibility to park features that its residents most care
about. In this paper, we proceed only incrementally beyond this theory by
adding Twitter activity as an element of a park's attractiveness above and
beyond its size. Exploring additional amenities or market segmentation strategies
could provide a more definitive understanding of the relationship between park
accessibility and health outcomes.

As a travel impedance measure, we used the Euclidean distance between each tract's
population-weighted centroid and the nearest point on the edge of a park's border.
Euclidean distances have well-rehearsed limitations regarding their fidelity with
the underlying infrastructure network, etc. Network-based distances can also
suffer from challenges when applied to multimodal problems; these challenges are
exacerbated when the non-highway mode share is high, as is likely when considering
access to parks. A better metric of travel impedance may be a mode choice model
logsum, which weights all travel alternatives against each other.

Finally, this study is primarily focused on the hypothesis that accessibility to
parks encourages physical activity, which in turn reduces obesity. There are a
multitude of other hypotheses that might be proposed and tested with the basic
methodology we have employed here, or competing explanations for the outcomes we
have observed. It is distinctly possible, for instance, that individuals who wish
to exercise regularly in parks choose to live near them. Given that the CDC models
generating the obesity and physical activity estimates presumably include variables
likely to influence such preferences (income, etc.), our investigation cannot
isolate the preferences from the effect. Controlling for such a
self-selection effect would be necessary to isolate the exogenous
impacts of park access on obesity or other health outcomes. And regarding these
other variables: this study did not consider potential relationships between park
access and hospitalization rates, life expectancy, respiratory disease,
mental health, or any number of potential beneficial outcomes hypothesized
or explored in the existing literature. Exploring these connections and their
underlying mechanisms should be a priority as city planners and urban architects
attempt to improve the quality of life of urban residents in the future.

\section{Conclusion}

Increased physical activity and decreased obesity rates are critical measures of
improvement in public health. Although many have theorized the link between park
accessibility and these metrics, previous literature has been mixed. Using New
York City as a case study, we presented a holistic and flexible measurement for
park accessibility that compares the continuous distance to all parks in the
region, weighted against the size of the park and its other amenities. In terms
of physical activity, we found a positive relationship where the least
park-accessible tracts have an expected physical activity participation rate
roughly one percentage point lower than the most accessible tracts. In terms of
obesity, we also found a strong and significant correlation between increased
park accessibility and decreased obesity rates, controlling for
spatial correlation, socioeconomic attributes, and physical activity rates.

In both cases a traditional, buffer-based analysis estimated a weaker relationship
with greater uncertainty. Buffer analyses are relatively common in the literature,
perhaps because of the widespread availability of GIS software. In spite of their
widespread use, they are relatively limited in terms of both their theoretical
underpinnings and their flexibility to accommodate attributes of parks beyond
their proximity. And even proximity may not be adequately handled, as the buffer
distance may be arbitrarily asserted by the researcher. Adopting choice-based
accessibilities of the kind used in this study will allow researchers to
encompass the full universe of park amenities in their accessibility analyses.
This will in turn enable planners to consider how multiple attributes of a park
--- from its location to its size to its amenities and beyond --- benefit the health
of the community the park serves.

\begin{funding}
This project was funded by the Speedwell Foundation.
\end{funding}
